---
title: "SME 10"
author: "Turki Bin Hammad"
date: "1/27/2022"
output: 
  pdf_document:
  highlight: pygments
header-includes:
    - \usepackage{setspace}
    - \doublespacing
number_sections: TRUE
---
\centering

# STATISTICAL METHODS IN EPIDEMIOLOGY

# SESSION 10: PRACTICAL

# Logistic regression 2


\raggedright



- Start a new R script and save it ("10_Rscript file" is a suggested name), install the required packages for the process, load their libraries and then read the *mortality data*, mortality.dta. You can use summarize, codebook, describe and structure to familiarize yourself with the data.

+  Start a new R script and save it

+ Install the required packages

```{r Requried packages, echo= T , eval=FALSE}

install.packages("haven")
install.packages("dplyr")
install.packages("epiDisplay")
install.packages("mStats")
install.packages("epitools")
install.packages("tidyverse")
install.packages("Epi")
install.packages("naniar")

```

-   Load the packages

```{r load packages, eval=T, echo=T, message=FALSE, warning=FALSE, paged.print=FALSE, results='hold'}

library("haven")
library("tidyverse")
library("epiDisplay")
library("dplyr")
library("tidyr")
library("mStats")
library("Epi")
library("biostat3")
library("rstatix")
library("insight")
library("naniar")
library("ggformula")
```


-   Read the dataset

```{r read dataset, eval=TRUE, echo=TRUE, results='hide'}
# We will use the read_dta() function from haven package 
#to read datasets saved as dta files
# You can check the details of this function by searching its name
#in the help tab or using question mark followed 
#by the function name ?read_dta()
mortality <- 
  read_dta("/Users/turkimohammed/Desktop/Study/SME/SME Datasets 2022/mortality.dta")     
# use the pathname from your computer
# Explore your dataset #
# Check the structure of your data
str(mortality)
# codebook is from mStats package 
codebook(mortality)
glimpse(mortality)
```


```{r, eval=TRUE, echo=TRUE, results='hide'}
#First we need to check that our outcome is coded as a factor with two levels
str(mortality$died)

# Transform it to factor
mortality$died <- factor(mortality$died, levels = c(1,0), labels = c("died","alive"))

#Do the same for exposure 
#Microfilarial infection  0: uninfected, 1:infected
str(mortality$mfpos)
mortality$mfpos <- factor(mortality$mfpos, levels = c(1,0), 
                        labels = c("infected","uninfected"))

# Tabulate the outcome by the exposure (using tab() function from mStats)
mortality %>% tab(died,by=mfpos)


# Using logistic regression to estimate crude OR

#Sometimes R will not choose the desired value 
#of the exposure or the outcome as a reference
#So we can set the reference values using relevel() function
mortality$mfpos <- relevel(mortality$mfpos, ref = "uninfected")
mortality$died <- relevel(mortality$died, ref = "alive")

#Fit the model
died_mfpos <- glm(died~mfpos, family =binomial(link = "logit") , data = mortality,
                 na.action = na.exclude)
# Get a summary
summary(died_mfpos)
# Calculate the OR
logistic.display(died_mfpos, decimal = 4)




```


```{r, eval=TRUE, echo=TRUE, results='hide'}
# Check agegrp 
str(mortality$agegrp)

# Transform to factor
mortality$agegrp <- factor(mortality$agegrp, levels = c(0,1,2,3), 
                           labels = c("15-34", "35-54", "55-64", "65+"))

# Tabulate using tab() function from mStats
mortality %>% drop_na() %>% tab(agegrp,by=mfpos)

```


```{r eval=TRUE, echo=TRUE, warning=FALSE, results='hide'}
# Some packages will have overlapping names of functions 
#so sometimes it is useful to detach the package and reload it to avoid error 
detach("package:epiDisplay", unload = TRUE)
library(epiDisplay)

##Mantel-Haenszel (M-H) Odds Ratios from epiDisplay
mhor(mortality$died,mortality$mfpos, strata = mortality$agegrp, decimal=4, graph = F,
design = "cohort")



```


```{r, eval=TRUE, echo=TRUE, results='hide'}

#Fit the model
dmfposage <- glm(died~mfpos+agegrp, family =binomial(link = "logit") , 
                 data = mortality,na.action = na.exclude)
#Get a summary
summary(dmfposage)
# Calculate ORs 
logistic.display(dmfposage, decimal = 4, crude = F)

```



```{r, eval=TRUE, echo=TRUE, results='hide'}
# Transform the vimp to factor 
mortality$vimp <- factor(mortality$vimp, levels = c(1,0), 
                         labels = c("Visually impaired","Normal"))

# Set the reference vimp 
mortality$vimp <- relevel(mortality$vimp, ref = "Normal")

#Fit the model 
morta_multi <- glm(died~vimp+mfpos+agegrp, family =binomial(link = "logit"), 
                   data = mortality,na.action = na.exclude)

# Get a summary
summary(morta_multi)

#calculate ORs 
logistic.display(morta_multi, decimal = 4, crude = F)


```


```{r, eval=TRUE, echo=TRUE, results='hide'}
# To perform a likelihood ratio test you need to fit a simpler model 
#and compare it to the full model
# In this case the full model includes vimp+mfpos+agegrp 
# Our simpler model will be without vimp while keeping mfpos+agegrp 
#since our assumption is that vimp does not have any effect on death 
#so will not produce any effect if we exclude it from the model

#This step was done in part 4 of the practical where we fitted 
#the odds ratio for onchocercal infection and age group "dmfposage"

# Conduct a likelihood ratio test (LRT)
lrtest(dmfposage,morta_multi)

```


```{r, eval=TRUE, echo=TRUE, results='hide'}
# In this step again we will have two models one with age group and the other without age group

# Fit a model without age group 
mortavimpmfpose <- glm(died~vimp+mfpos, family =binomial(link = "logit"), 
                       data = mortality,na.action = na.exclude)

#The model with age group is "morta_multi"

# Conduct a likelihood ratio test (LRT)
lrtest(mortavimpmfpose,morta_multi )


```



- Read the dataset 
```{r, eval=TRUE, echo=TRUE, results='hide'}
Mwanza <- 
  read_dta("/Users/turkimohammed/Desktop/Study/SME/SME Datasets 2022/mwanza.DTA")  
# use the pathname from your computer
# Explore your dataset #
# Check the structure of your data
str(Mwanza)
# codebook is from mStats package 
codebook(Mwanza)
glimpse(Mwanza)

```



```{r, eval=TRUE, echo=TRUE, results='hide', fig.keep='none'}
# We will create a new variable using case_when()
Mwanza <- Mwanza %>% mutate( # Create categories 
     ed2 = as.factor(case_when(
      ed == 1 ~ "none/adult only",
      ed >= 2 ~ "1 or more years of formal education"
    )))
# You can label the variable using var_label() from labelled package
# To check that you have labeled the variable correctly
#you can used the codebook  from mStats package again
var_label(Mwanza$ed2) <- "Education"
codebook(Mwanza)
# After creating a new variable, it is good to tabulate
#it to ensure that your steps was accurate and no mistakes happened
#We will use tab1 fro epiDisplay package
tab1(Mwanza$ed2)

```


```{r, eval=TRUE, echo=TRUE, results='hide'}
#Since our cases were read as numerical variable we will need
#to transform it to a factor using the following code
#to be able to tabulate it
Mwanza$case <- factor(Mwanza$case, levels = c(1,0), labels = c("Yes", "No")) 


# To create our 2x2 table we will use tab(#cases, by= #exposure)
#function from mStats package
Mwanza %>% tab(case,by=ed2, na.rm = F)

```



```{r, eval=TRUE, echo=TRUE, results='hide',fig.keep='none'}
#Tabulate the variable 
tab1(Mwanza$rel)

# Assign 9=missing to NA
Mwanza<- Mwanza %>% replace_with_na(replace = list(rel = 9))

# Get the sum of NAs 
sum(is.na(Mwanza$rel))

#Tabulate the variable again  
tab1(Mwanza$rel)


#Transform religion to a factor
#religion: 1=Muslim 2=Catholic 3=Protestant 4=other 
Mwanza$rel <- factor(Mwanza$rel, levels= c(1,2,3,4), 
            labels = c("Muslim", "Catholic", "Protestant","other"))
```



```{r, eval=TRUE, echo=TRUE, results='hide'}
library(epiDisplay)
##Mantel-Haenszel (M-H) Odds Ratios from epiDisplay
mhor(Mwanza$case,Mwanza$ed2, strata = Mwanza$rel, decimal=4, graph = F,
design = "case-control")

```


```{r, eval=TRUE, echo=TRUE, results='hide'}
#Fit the model odds ratio for ed2
oddsed2 <- glm(case~ed2, family =binomial(link = "logit"),
               data = Mwanza,na.action = na.exclude)
# Get a summary
summary(oddsed2)


#Fit the model odds ratio for ed2controlling religion
oddsed2rel <- glm(case~ed2+rel, family =binomial(link = "logit"),
                  data = Mwanza,na.action = na.exclude)
# Get a summary
summary(oddsed2rel)

#calculate ORs 
#This function will give you the crude versus adjusted
logistic.display(oddsed2rel, decimal = 4, crude = T)

#Another option is to use the function again for oddsed2
logistic.display(oddsed2, decimal = 4, crude = T)


```

```{r, eval=TRUE, echo=TRUE, results='hide'}
# In this step  we will have two models one with ed2 and the other without ed2

# Fit a model without age group 
oddsrel <- glm(case~rel, family =binomial(link = "logit"), 
               data = Mwanza,na.action = na.exclude)

#The model with ed2 is "oddsed2rel"

# Conduct a likelihood ratio test (LRT)
lrtest(oddsed2rel,oddsrel)


```

