---
title: 'SME SESSION 9: PRACTICAL'
author: "Turki Bin Hammad"
date: "1/21/2022"
output: 
  pdf_document:
  highlight: pygments
header-includes:
    - \usepackage{setspace}
    - \doublespacing
number_sections: TRUE
---

\centering

# STATISTICAL METHODS IN EPIDEMIOLOGY

# SESSION 9: PRACTICAL

# Logistic regression 1

\raggedright

1.  Start a new R script and save it ("9_Rscript file" is a suggested name), install the required packages for the process, load their libraries and then read the *mortality data*, mortality.dta. You can use summarize, codebook, describe and structure to familiarize yourself with the data.

-   Start a new R script and save it

-   Install the required packages

```{r Requried packages, eval=FALSE, include=FALSE}

install.packages("haven")
install.packages("dplyr")
install.packages("epiDisplay")
install.packages("mStats")
install.packages("epitools")
install.packages("tidyverse")
install.packages("Epi")
```

-   Load the packages

```{r load packages, echo=FALSE, message=FALSE, warning=TRUE}

library("haven")
library("tidyverse")
library("epiDisplay")
library("dplyr")
library("tidyr")
library("mStats")
library("Epi")
library("biostat3")
library("rstatix")
library("insight")

```


-   Read the dataset

```{r read dataset, eval=TRUE, echo=TRUE, results='hide'}
# We will use the read_dta() function from haven package 
#to read datasets saved as dta files
# You can check the details of this function by searching its name
#in the help tab or using question mark followed 
#by the function name ?read_dta()
mortality <- 
  read_dta("/Users/turkimohammed/Desktop/Study/SME/SME Datasets 2022/mortality.dta")     
# use the pathname from your computer
# Explore your dataset #
# Check the structure of your data
str(mortality)
# codebook is from mStats package 
codebook(mortality)
glimpse(mortality)
```


```{r, eval=F, echo=TRUE, results='hide'}
#First we need to check that our outcome is coded as a factor with two levels
str(mortality$died)

# Transform it to factor
mortality$died <- factor(mortality$died, levels = c(1,0), labels = c("died","alive"))

#Tabulate it to ensure everything was correct (using tab1() function from epiDisplay)
tab1(mortality$died)

# Tabulate the outcome by the exposure (using tab() function from mStats)
mortality %>% tab(died,by=vimp)


#We will convert our exposure variable vimp to factor 
mortality$vimp <- factor(mortality$vimp, levels = c(1,0), 
                         labels = c("Visually impaired","Normal"))

#Calculate the odds in each group and the crude odds ratio (OR) using mhor from mStats 
# To avoid having the reference levels confused 
#we will define it using exp_value = 1, case_value = 1

mhor(mortality,
  exp=vimp ,
  case=died,
  strata = NULL,
  exp_value = 1,
  case_value = 1,
  digits = 4)


```


```{r, eval=F, echo=TRUE, results='hide'}

#Sometimes R will not choose the desired value of the exposure or the outcome as a reference
#So we can set the reference values using relevel() function
mortality$vimp <- relevel(mortality$vimp, ref = "Normal")
mortality$died <- relevel(mortality$died, ref = "alive")


#Compute a logistic regression model 

#To fit a logistic regression model in R you need to use the glm() function

#This function works as follows 
#glm(outcome~exposure, then you the likelihood distribution, 
#specify your data, 
#specify how you will deal with missing values(i.e. NAs))
# Since we wanted to work on a log binomial scale we chose binomial(link = "logit")
#There are different options as most of the regressions in R will use the same function 
# glm() but the difference will be in the family = .... 

mortaregg <- glm(died~vimp, family =binomial(link = "logit") , data = mortality,
                 na.action = na.exclude)

mortaregg

#We can get more details about our model 
summary(mortaregg)

# Calculate the log likelihood of the model 
get_loglikelihood(mortaregg)


#Note that the estimates you have are on log scale and not the normal scale  
```


```{r, eval=F, echo=TRUE, results='hide'}
#Calculate odds ratio from the model 
#using logistic.display() function from epiDisplay package

logistic.display(mortaregg, alpha = 0.05, crude = TRUE, 
    crude.p.value = FALSE, decimal = 4, simplified = FALSE)

# This function does not show the Intercept (_cons in STATA)

# OR
exp(coef(mortaregg),confint.default(mortaregg))
#95% CI 

# You can conduct a wald test  (this is another option) and 
#a Likelihood ratio test (LRT) using the functions below

waldtest(mortaregg)
lrtest(mortaregg)


```



```{r, eval=F, echo=TRUE, results='hide'}
#We will convert our exposure variable mfgrp to factor 
str(mortality$mfgrp)

mortality$mfgrp <- factor(mortality$mfgrp, levels = c(0,1,2,3), 
                          labels = c("Uninfected ", "<10", "10-49", "50+"))

#Tabulate it to make sure that everything went smooth 
tab1(mortality$mfgrp)

# Tabulate the outcome by the exposure # We choose to remove NAs 
mortality %>% tab(died,by=mfgrp, na.rm= T)



#Calculate the odds in each group and the crude odds ratio (OR) using mhor from mStats 
# To avoid having the reference level confused we will define it using exp_value = 

mhor(mortality,
  exp=mfgrp ,
  case=died,
  strata = NULL,
  exp_value = NULL,
  case_value = 1,
  digits = 4)



```


```{r, eval=F, echo=TRUE, results='hide'}
# Before fitting our model, it is always important that you understand your data
#especially the missing data 

# To examine the number missingness in our exposure and outcome we can use is.na()
sum(is.na(mortality$mfgrp))
sum(is.na(mortality$died))

# Since our exposure variable has 93 missing value and our outcome has 0 
# We can create a new variable for our outcome which has the value NA if the exposure is NA
#This step is important in further steps especially LRT 

#We created a new variable called died1 which equals died with value NA if the exposure is NA  
library(data.table)
setDT(mortality)[!is.na(mfgrp) , died1 := died]

#We can recheck how many NAs we have in each variable 
sum(is.na(mortality$mfgrp))
sum(is.na(mortality$died1))



#Compute a logistic regression model #We will use died1 rather than died!
mortamfgrp <- glm(died1~mfgrp, family =binomial(link = "logit") , data = mortality,
                  na.action = na.exclude)
mortamfgrp
summary(mortamfgrp)


# Calculate the log likelihood of the model 
get_loglikelihood(mortamfgrp)


#Calculate odds ratio from the model 
logistic.display(mortamfgrp, decimal = 4)


# You can conduct a Likelihood ratio test (LRT) using the functions below
lrtest(mortamfgrp)

```



```{r, eval=F, echo=TRUE, results='hide'}
#Although we removed the NAs values in the outcome previously 
#we can explore this issue in following example 


#First we will Compute another logistic regression model mortadied
#We will use died rather than died1 this time 

mortadied <- glm(died~1, family =binomial(link = "logit") , data = mortality,
                 na.action = na.exclude)

#Then we will Compute another logistic regression model mortadied1
#We will use died1 rather than died this time 
#Remember that we assign the NAs value in the exposure as NAs in the outcome  

mortadied1 <- glm(died1~1, family =binomial(link = "logit") , data = mortality,
                  na.action = na.exclude)

# We can conduct a Likelihood ratio test (LRT) now 

#Using mortadied1 compared with our previous model mortamfgrp 
lrtest(mortadied,mortamfgrp)
#You will get an error message saying models were not all fitted to the same size of dataset

#Now do that using mortadied1 compared with our previous model mortamfgrp 
lrtest(mortadied1,mortamfgrp)

###This should work ###



```


```{r, eval=F, echo=TRUE, results='hide'}
#We will convert our exposure variable agegrp to factor 
str(mortality$agegrp)

mortality$agegrp <- factor(mortality$agegrp, levels = c(0,1,2,3), 
                           labels = c("15-34", "35-54", "55-64", "65+"))

#Tabulate it to make sure that everything went smooth 
tab1(mortality$agegrp)

# Tabulate the outcome by the exposure # We choose to remove NAs 
mortality %>% tab(died,by=agegrp, na.rm= T)


#Compute a logistic regression model #We will use died1 rather than died!
mortagegrp <- glm(died~agegrp, family =binomial(link = "logit") , data = mortality,
                  na.action = na.exclude)
mortagegrp
summary(mortagegrp)


# Calculate the log likelihood of the model 
get_loglikelihood(mortagegrp)


#Calculate odds ratio from the model 
logistic.display(mortagegrp, decimal = 4)


# You can conduct a Likelihood ratio test (LRT) using the functions below
lrtest(mortagegrp)

```


```{r, eval=F, echo=TRUE, results='hide'}

#Compute a logistic regression model #We will use died1 rather than died!
mortacombined <- glm(died~ agegrp+vimp, family =binomial(link = "logit") , data = mortality,
                     na.action = na.exclude)
mortacombined
summary(mortagegrp)


# Calculate the log likelihood of the model 
get_loglikelihood(mortagegrp)


#Calculate odds ratio from the model 
logistic.display(mortacombined, decimal = 4,crude = FALSE )


# You can conduct a Likelihood ratio test (LRT) using the functions below
lrtest(mortagegrp)

```
